Regression Cube Analysis of Cohort Study Data

### 2015-06-21 draft.tex
cut several parts to shorten the paper

We designed our system to be openly accessible and easy to use.

@@

The design choice of using web-based technologies offers various advantages w.r.t. the collaboration with epidemiologists.

@@

%% INTERACTION
Rotating with the cube is restricted to the y-axis, preserving the mental map to position individual features.
%%
The cube is always oriented according to the 2D representation, allowing for an easy mental combination of the two representations.
%%
Allowing more degrees of freedom was confusing to our users and also did not add value to the visualization.
%%
Also we provide a zoom functionality using the mouse wheel input.
%%
\paragraph{Cube Slice Selection [3.ZF].}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate to different slices of interest.
%%
We propose two ways to achieve this.
\begin{itemize}
	\item \emph{Applying the slicing metaphor from 3D volume data.}
	%%
	In medical volume data renderings, slicing views are very common to view details on a selected plane in the scene.
	%%
	We employ this technique for selecting cube slices (e.g., by moving a plane via vertical mouse input while pressing the right mouse button).
	%%
	We, however, still display the whole 3D object instead of cutting away information down to the slice position.
	%%
	\design{Early prototypes only provided this method to select a slice of interest, which was inefficient when the user was looking for a specific slice.
	%%
	Hence, an additional method was implemented.
	%%
	}
	%%
	\item \emph{Selecting the slice using a dropdown menu} provides fast access to plane selections when the user already knows the slices of interest.
	%%
\end{itemize}
%%
The currently selected slice is denoted with a semi-transparent gray plane.
%%
\design{
Early designs rendered the whole cube, which made it hard to assess the position of the plane.
%%
%Since the regression metrics are mirrored along the diagonal, we decided to use this space to support better identification of the current slice.
Since the regression metrics are mirrored along the diagonal, the space available from visualizing only the prism generated from the upper half of the cube diagonal is used to display information about the currently selected plane.
}
%%
The space available from visualizing only the prism generated from the upper half of the cube diagonal is used to display information about the currently selected plane.
%%
The \magenta{regression metric values} are projected on this plane to provide an overlapping-free view.
%%
This allows for a easier to identify the current slice.
%%
\paragraph{2D Mosaic Plot Slice Visualization [4.DD].}
%%
The 2D mosaic visualization (Fig.~\ref{fig:System}c) shows all values below the matrix diagonal of the current slice.
%%
It creates a optical equivalence with the 3D cube.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation-based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D cube.
%%

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server back-end (right) is written using \emph{Node.js}, hosting is carried out using \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{Regression Cubes}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
 %%
 }
  \label{fig:Implementation}
\end{figure}
%%
We rely on web-based technologies for our prototype.
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
It forces developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of user interface elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents} (\texttt{D3.js}), which is popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
%%
Open source volume rendering methods are available but do not satisfy our requirements.
%%
Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
%%
Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
%%
It also is the fastest solution that we tested.

\paragraph{Back-End.}
Two server structures serve as back-end.
%%
The first one is the web-server, which is written in \texttt{Javascript} using \texttt{Node.js}\footnote{Open Source; Maintained by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, running on Googles V8 Javascript runtime environment.
%%
It is hosted on \texttt{Heroku}\footnote{Owned by Salesforce.com, \href{https://www.heroku.com/}{\texttt{heroku.com}}}, a cloud application platform.
%%

The statistical processors yield the second structure.
%%
They rely on the statistical programming language \texttt{R}.\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is an \texttt{R} package and provides an API for accessing it via HTTP calls \cite{Ooms}.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
The back-end functions necessary for all cube calculations are provided via an \texttt{R} package.
%%
It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
%%
The server workload balances are managed by the front-end code.

\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}.\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\\\texttt{github.com/paulklemm/regression-cube-r-package}}}$^{,}$\footnote{Front-End and Node.js Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{\\github.com/paulklemm/regression-cube-prototype}}}
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical back-end through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku app.
%%
\section{Application} \label{application}
%%
\begin{figure*}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/application}
 \caption{
 The analysis of the numerical and dichotomized target feature depicting liver fat values yields similar results (left).
 %%
 In (a), hotspots for somatometrical features with high correlations were found.
 %%
 High correlations were also found for features depicting \emph{hepatic steatosis} (b).
 %%
 A high correlation between \emph{Interleukin-6}, \emph{hepatic steatosis}, \emph{GGT} and \emph{Lipase} (highlighted using arrows) was revealed during the analysis using the 2D mosaic plot.
 %%
 The hypothesis-free analysis of the breast fat data set (right) w.r.t. the \emph{parenchyma tissue percentage} of the breast displays correlations between \emph{age}, \emph{body fat}, \emph{hip circumference} as well as \emph{menstrual period}.
 }
  \label{fig:Application}
\end{figure*}
%%
In this section, we describe how we applied the \emph{Regression Cube} to two epidemiological data sets.
%%
The hepatic steatosis data set was analyzed using data mining algorithms, yielding risk groups, which we now analyze further.
%%
Also, we try to reproduce the prior results from this analysis as proof-of-concept of our method.
%%
The breast fat data set is the basis for an explorative analysis w.r.t. the influencing parameters of the parenchyma tissue of the female breast.
%%

>>>>

\noindent We designed our system to be openly accessible and easy to use.

@@

Using web-based technologies offers various advantages w.r.t. the collaboration with epidemiologists.

@@

%% INTERACTION
Rotating with the cube is restricted to the y-axis, preserving the mental map to position individual features.
%%
The cube is always oriented according to the 2D representation, allowing for an easy mental combination of the two representations.
%%
Allowing more degrees of freedom was confusing to our users and also did not add value to the visualization.
%%
Also we provide a zoom functionality using the mouse wheel input.
%%
\paragraph{Cube Slice Selection [3.ZF].}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate to different slices of interest.
%%
We propose two ways to achieve this.

%\begin{itemize}
	%\item 
	\textbf{Applying the slicing metaphor from 3D volume data.}
	%%
	In medical volume data renderings, slicing views are very common to view details on a selected plane in the scene.
	%%
	We employ this technique for selecting cube slices (e.g., by moving a plane via vertical mouse input while pressing the right mouse button).
	%%
	We, however, still display the whole 3D object instead of cutting away information down to the slice position.
	%%
	\design{Early prototypes only provided this method to select a slice of interest, which was inefficient when the user was looking for a specific slice.
	%%
	Hence, an additional method was implemented.
	%%
	}
	%%
	%\item

	\textbf{Selecting the slice using a dropdown menu} provides fast access to plane selections when the user already knows the slices of interest.
	%%
%\end{itemize}
%%
The currently selected slice is denoted with a semi-transparent gray plane.
%%
\design{
Early designs rendered the whole cube, which made it hard to assess the position of the plane.
%%
%Since the regression metrics are mirrored along the diagonal, we decided to use this space to support better identification of the current slice.
Since the regression metrics are mirrored along the diagonal, the space available from visualizing only the prism generated from the upper half of the cube diagonal is used to display information about the currently selected plane.
}
%%
The space available from visualizing only the prism generated from the upper half of the cube diagonal is used to display information about the currently selected plane.
%%
The \magenta{regression metric values} are projected on this plane to provide an overlapping-free view.
%%
This allows for a easier to identify the current slice.
%%
\paragraph{2D Mosaic Plot Slice Visualization [4.DD].}
%%
The 2D mosaic visualization (Fig.~\ref{fig:System}c) shows all values below the matrix diagonal of the current slice.
%%
It creates a optical equivalence with the 3D cube.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation-based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D cube.
%%

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server back-end (right) is written using \emph{Node.js}, hosting is carried out using \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{Regression Cubes}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
 %%
 }
  \label{fig:Implementation}
\end{figure}
%%
We rely on web-based technologies for our prototype.
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.

%\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
It forces developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of user interface elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents} (\texttt{D3.js}), which is popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
% We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
% %%
% Open source volume rendering methods are available but do not satisfy our requirements.
% %%
% Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
% %%
% Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
% %%
% It also is the fastest solution that we tested.

%\paragraph{Back-End.}
Two server structures serve as back-end.
%%
The first one is the web-server, which is written in \texttt{Javascript} using \texttt{Node.js}\footnote{Open Source; Maintained by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, running on Googles V8 Javascript runtime environment.
%%
It is hosted on \texttt{Heroku}\footnote{Owned by Salesforce.com, \href{https://www.heroku.com/}{\texttt{heroku.com}}}, a cloud application platform.
%%

The statistical processors yield the second structure.
%%
They rely on the statistical programming language \texttt{R}.\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is an \texttt{R} package and provides an API for accessing it via HTTP calls \cite{Ooms}.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
% The back-end functions necessary for all cube calculations are provided via an \texttt{R} package.
% %%
% It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
% %%
% The server workload balances are managed by the front-end code.

%\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}.\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\\\texttt{github.com/paulklemm/regression-cube-r-package}}}$^{,}$\footnote{Front-End and Node.js Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{\\github.com/paulklemm/regression-cube-prototype}}}
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical back-end through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku app.
%%
\section{Application} \label{application}
%%
\begin{figure*}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/application}
 \caption{
 The analysis of the numerical and dichotomized target feature depicting liver fat values yields similar results (left).
 %%
 In (a), hotspots for somatometrical features with high correlations were found.
 %%
 High correlations were also found for features depicting \emph{hepatic steatosis} (b).
 %%
 A high correlation between \emph{Interleukin-6}, \emph{hepatic steatosis}, \emph{GGT} and \emph{Lipase} (highlighted using arrows) was revealed during the analysis using the 2D mosaic plot.
 %%
 The hypothesis-free analysis of the breast fat data set (right) w.r.t. the \emph{parenchyma tissue percentage} of the breast displays correlations between \emph{age}, \emph{body fat}, \emph{hip circumference} as well as \emph{menstrual period}.
 }
  \label{fig:Application}
\end{figure*}
%%
In this section, we describe how we applied the \emph{Regression Cube} to two epidemiological data sets.
%%
The hepatic steatosis data set was analyzed using data mining algorithms, yielding risk groups, which we now analyze further.
%%
Also, we try to reproduce the prior results from this analysis as proof-of-concept of our method.
%%
The breast fat data set is the basis for an explorative analysis w.r.t. the influencing parameters of the parenchyma tissue of the female breast.
%%
